Ionization of biological molecules by multicharged ions by using the stoichiometric model

### 2019-09-26 NewMonti25.tex
checked with grammarly and new figs

%\documentclass[pra,showpacs,twocolumn]{revtex4}
\documentclass[preprint]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{gensymb}
\usepackage{float}
\usepackage{authblk}
\usepackage{MnSymbol}
\usepackage{gensymb}

@@

\author{author}
%\affiliation{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio 
%(CONICET-UBA).
%Casilla de correo 67, sucursal 28 (C1428EGA) Buenos Aires, Argentina.}
%\author{author}
%\affil{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio (CONICET-UBA).
%Casilla de correo 67, sucursal 28 (C1428EGA) Buenos Aires, Argentina.}
%\affil{Fac. de Ciencias Exactas y Naturales, Universidad de Buenos Aires}

\date{\today}% It is always \today, today,

\begin{abstract}
\end{abstract}

@@

To overcome the first perturbative approximation limitations, and since 
the projectiles are multicharged ions, we resort to the Continuum 
Distorted Wave--Eikonal Initial State (CDW), which includes higher 
perturbative corrections. We start from the premise that the ionization 
process is the mechanism that deposits the most significant amount of 
primary energy. Moreover, it is well known that the residual electrons 
from the ionization become a source of significant local biological 
damage. The secondary electrons are included in Monte Carlo simulations, 
and hence their behavior must be investigated. 
In Section~\ref{subsec:meanener} and \ref{subsec:meanang}, we calculate
the mean energy and angular distributions of the ejected electrons. 
Surprisingly, we found a substantial dependence of the charged 
projectile, which is unexpected in the first Born approximation. 
We deal with the molecular structure complexity of the target by 
implementing the simplest stoichiometric model (SSM): the molecules are 
assumed to be composed of isolated independent atoms, and the total 
cross--section by a linear combination of weighted atomic calculations.

By implementing the CDW and the SSM, we calculate ionization 
cross--section of several molecules of interest (see 
Table~\ref{tab:families}) by the impact of antiprotons, H$^{+}$, 
He$^{+2}$, Be$^{+4}$, C$^{+6}$, and O$^{+8}$. 
In Section~3, we show our results for different DNA and RNA molecules, 
such as adenine, cytosine, guanine, thymine, uracil, tetrahydrofuran~(THF), 
pyrimidine, and also DNA backbone. In Section~\ref{subsec:scaling}, we
test the Toburen scaling rule~\cite{toburen1975,toburen1976}, which 
states that the ratio between the ionization cross--section and the 
number of weakly bound electrons can be arranged in a narrow universal 
band in terms of the projectile velocity. We applied this rule to 
several hydrocarbons and nucleobases and noted that the width of the 
resulting universal band could be significantly reduced if we redefine 
the effective number of active electrons in the collision. The new 
scaling is then tested by comparison with experimental data.

Although the multicharged projectiles are dealt with the CDW, the 
stoichiometric model used seems to be very simplistic. The approach
considers each atom as neutral, which is not correct. In Section 3.5, 
we used the molecular electronic structure code 
{\sc gamess}~\cite{gamess} to calculate the excess or defect of 
electron density on the atoms composing the molecules. Then, the SSM is 
modified to account for the departure from the neutrality of the atoms. 
We find that the modified SSM for the DNA molecules does not introduce 
substantial changes in the cross sections.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory: Ionization of Atoms}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In the present study, we consider six atoms, $\alpha=$ H, C, N, O, P, 
and S, and six projectiles, antiprotons $\bar{p}$, H$^{+}$, He$^{+2}$, 
Be$^{+4}$, C$^{+6}$, and O$^{+8}$. 
Most of the organic molecules are composed of these atoms. Some 
particular molecules also include halogen atoms such as fluor and 
bromine; ionization cross sections of these elements have been 
previously published~\cite{miraglia2008,miraglia2009}.

The total ionization cross sections of these atoms were calculated using 
the CDW~[xxx]. The initial bound and final continuum radial wave 
functions were obtained by using the {\sc radialf} code, developed by 
Salvat and co--workers~\cite{salvat1995}, and a Hartree-Fock potential 
obtained from the Depurated Inversion Method~\cite{mendez2016,mendez2018}. 
We used a few thousand pivot points to solve the Schr\"{o}dinger 
equation, depending on the number of oscillations of the continuum 
state. The radial integration was performed using the cubic spline 
technique. The number of angular momenta considered varied from 8, at 
very low ejected-electron energies, up to 30, for the highest energies 
considered. The same number of azimuth angles were required to obtain 
the four-fold differential cross section. The calculation performed does 
not display prior-post discrepancies at all. Each atomic total cross 
section was calculated using 35 to 100 momentum transfer values, 28 
fixed electron angles, and around 45 electron energies depending on the 
projectile impact energy. In our theoretical treatment, we expand our 
final continuum wave function as usual,
\begin{equation}
\psi_{\overrightarrow{k}}^{-}(\overrightarrow{r})=\sum_{l=0}^{l_{\max
}}\sum_{m=-l}^{l}R_{kl}^{-}(r)Y_{l}^{m}(\widehat{r})Y_{l}^{m^{\ast }}
(\widehat{k})\,.
\label{eq:contwave}
\end{equation}
We are confident with our calculations up to $l_{\max}\sim 30$. Further 
details of the calculation are given in Ref.~\cite{montanari2017}. 
Simultaneously, we will be reporting state to state ionization cross 
sections for the 36 ion--target systems considered in the present 
work~\cite{miraglia2019}. We expect that these results will be useful
to estimate molecule fragmentation.

\begin{figure*}[t!]
\centering
%\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/Fig1.eps}
\includegraphics[width=0.9\textwidth]{figuras/atomicscaling.eps}
\caption{Reduced CDW total ionization cross section of six atomic 
targets. The curves are labeled with the charge state corresponding to 
the six multicharged projectiles.}
\label{fig:atomscaling}
\end{figure*}

We report our total CDW ionization cross sections for the six essential 
elements by the impact of six projectiles in Fig.~\ref{fig:atomscaling}.
To reduce the resulting 36 magnitudes into a single consistent 
figure, we considered the fact that in the first Born approximation
the ionization cross section scales with the square of the projectile 
charge, $Z^{2}$. The values of the impact energies considered 
range between 0.1 to 10 MeV/amu, where the CDW is supposed 
to hold. In fact, for the highest projectile charges the minimum 
impact energy where the CDW is expected to be valid could be 
higher than 100 keV. We also performed similar calculations with the 
first Born approximation, and we corroborated that it provides quite 
reliable results only for energies higher than a couple of MeV/amu. 
We use the same line color to indicate the projectile charge throughout 
all the figures of this work: dashed--red, solid--red, blue, magenta, 
olive and orange for antiprotons, H$^{+}$, He$^{+2}$, Be$^{+4}$, 
C$^{+6}$, and O$^{+8}$, respectively. Notably, there is no complete 
tabulation of ionization of atoms by the impact of multicharged ions. 
We hope that the ones presented in this article will be of help for 
future works.

@@

\begin{figure*}[t!]
\centering
%\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/fig5.eps}
\includegraphics[width=0.9\textwidth]{figuras/ener_mean.eps}
\caption{Mean emitted energy distribution for ionization by impact of
multicharged ions, given by Eq.~(\ref{40}). Dashed lines for $\bar{p}$,
solid lines for ion charges $+1$, $+2$, $+4$, $+6$ and $+8$,
as indicated.}
\label{fig:emittedener}
\end{figure*}

@@

In Fig.~\ref{fig:emittedener}, we estimate $\overline{E}_{\alpha}$ of
the emitted electron in the 0.5--2.7 a.u. energy range, or equivalently 
from 15 to 70 eV, for all the targets. Our results agree with the 
experimental findings~\cite{surdutovic2018}. As can be noted in the 
figure, the mean energy value is surprisingly sensible to the 
projectile charge $Z$, which can duplicate the proton results in the 
intermediate region, i.e. 100--400 keV/amu. The effect observed can be 
attributed to the depletion caused by the multicharged ions to the 
yields of low energy electrons. This behavior cannot be found in the 
first Born approximation, where the $Z^2$ law cancels the $Z$ dependence
in Eq.~(\ref{40}). At high energies, $\overline{E}_{\alpha}$ tends to a 
universal value for all ions, as can be seen in Fig.~\ref{fig:emittedener}.

@@

As mentioned before, secondary electrons contribute to the total damage. 
Then, not only is the ejection energy important but also the angle 
of emission. Once again, we can consider the single differential cross 
section in terms of the ejected electron solid angle $\Omega$, 
$d\sigma_{\alpha,nl}/d\Omega$, to be expressed as a distribution function, 
and the mean angle $\overline{\theta}_{\alpha}$ can be defined as
\begin{equation}
\overline{\theta}_{\alpha}=\frac{\langle\theta_{\alpha}\rangle}
{\langle 1\rangle}=\frac{1}{\sigma_{\alpha}}\sum\limits_{nl}
\int d\Omega\,\theta\,\frac{d\sigma_{\alpha,nl}}{d\Omega}
\end{equation}

\begin{figure*}[t!]
\centering
%\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/fig6.eps}
\includegraphics[width=0.9\textwidth]{figuras/ang_mean.eps}
\caption{Mean emitted angle distribution for ionization by impact of
multicharged ions.}
\label{fig:emittedang}
\end{figure*}

The mean emitted electron angles $\overline{\theta}_{\alpha}$ for the 
six atoms of interest are shown in Fig.~\ref{fig:emittedang}, and an 
important dependence of $\overline{\theta}_{\alpha}$ with $Z$ is 
observed. Once again, this effect is not observed in the first Born 
approximation. It is a general belief~\cite{Rudd1992} that the 
angular dispersion of emitted electrons are nearly isotropic. This 
behavior is caused by the insignificant angular anisotropy of 
sub--50--eV yield. 
A typical value for the ejection angle considered in the literature is 
$\overline{\theta}_{\alpha}\sim 70\degree$~\cite{surdutovic2018}, and 
it is quite correct in the range of validity of the first Born 
approximation for any target. But, when a distorted wave approximation 
is used, $\overline{\theta}_{\alpha}$ decreases substantially with $Z$ 
in the intermediate energy region, as observed in Fig.~\ref{fig:emittedang}. 
For example, for C$^{+6}$ impact, the Bragg peak occurs at 0.3 MeV/amu, 
where $\overline{\theta}_{\alpha}$, computed with the CDW method, is 
about half of the value obtained with the first Born approximation. 
This correction closes the damage to the forward direction.

We can attribute this forward direction correction to the capture to 
the continuum effect; the larger the charge $Z$, the smaller 
$\overline{\theta}$ will be. Of course, this effect only holds at 
intermediate energies and not at high impact energies, where the Born 
approximation rules. One illustrative observation is the behavior of 
antiprotons: the projectile in this case repels the electrons, being
$\overline{\theta}_{\alpha}\sim 90\degree$.
Note the opposite effect of proton and antiprotons; they run one 
opposite to the other, as compared with the first Born approximation,
constituting an angular Barkas effect.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Ionization of Molecules}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The stoichiometric model}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Lets us consider a molecule $M$ composed by $n_{\alpha}$ atoms of the
element $\alpha$, the SSM approaches the total ionization cross section 
of the molecule $\sigma_{M}$ as a sum of ionization cross sections of 
the isolated atoms $\sigma_{\alpha}$ weighted by $n_{\alpha}$, 
\begin{equation}
 \sigma_{M}=\sum\limits_{\alpha}n_{\alpha}\sigma_{\alpha}\,.  
 \label{eq:sumion}
\end{equation}
We classified the molecular targets of our interest in three families: 
CH, CHN and DNA, as in Table~\ref{tab:families}.

@@

In Fig.~\ref{fig:crossDNA_1}, we report the total ionization cross 
sections by the impact of multicharged ions for adenine, cytosine, 
guanine and thymine. For adenine, the agreement with the experimental 
data available~\cite{iriki2011} is very good. To the best of our 
knowledge, there are not experimental data on ion--collision ionization 
for the rest of the molecules. We included electron impact 
measurements~\cite{rahman2016} with the corresponding equivelocity 
conversion for incident energies higher than 300~eV. In this region, 
the proton and electron cross section should converge. Although the 
electron impact measurements are above our findings for all the 
molecular targets, it is worth stating that our results agree very well 
with other electron impact theoretical 
predictions~\cite{mozejko2003,tan2018}.

\begin{figure*}[t!]
\centering
%\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/fig2.eps}
\includegraphics[width=0.9\textwidth]{figuras/adn1.eps}
\caption{Reduced CDW ionization cross section by impact of multicharged 
ions. Experiments: \mbox{\Large$\circ$}~\cite{iriki2011} for proton 
impact and $\square$~\cite{rahman2016} for electron impact with 
equivelocity conversion.}
\label{fig:crossDNA_1}
\end{figure*}

Our results for uracil, DNA backbone, pyrimidine and THF are displayed 
in Fig.~\ref{fig:crossDNA_2}. For uracil, the agreement with the 
experimental proton impact measurements by 
Itoh~{\it et al.}~\cite{itoh2013} is good.
However, for the same target, our theory fails by a factor of two for 
the experimental ionization values by the impact of 
C$^{+4}$, O$^{+6}$ and O$^{+8}$ ions~\cite{agnihotri2012,agnihotri2013}.
Nonetheless, it should be stated that our theoretical results coincide 
with calculations by Champion, Rivarola and 
collaborators~\cite{agnihotri2012,champion2012}, which may indicate a 
possible misstep of the experiments. 
For pyrimidine, we show comparison of our results with experimental data
for proton~\cite{wolff2014} and electron~\cite{bug2017} ionization. 
The electron impact measurements 
agree with our calculations for energies higher than 500keV. 
Unexpectedly, the proton impact cross sections are significantly lower 
than our findings. 
The THF molecule results are compared with proton~\cite{wang2016} 
and electron~\cite{bug2017,wolf2019,fuss2009} impact measurements, showing
overall good agreement with our theory.

\begin{figure*}[t!]
\centering
%\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/fig3.eps}
\includegraphics[width=0.9\textwidth]{figuras/adn2.eps}
\caption{Reduced CDW ionization cross section by impact of multicharged 
ions. Experiments: proton impact on $\triangle$ uracil~\cite{itoh2013}, 
$\bigtriangledown$ pyrimidine~\cite{wolff2014} and $\meddiamond$
THF~\cite{wang2016}. Impact of $\ominus$ C$^{+4}$, 
$\boxminus$ C$^{+6}$, O$^{+6}$, F$^{+6}$, and
$\triangle\mkern-14mu-$ O$^{+8}$, F$^{+8}$ on 
uracil~\cite{agnihotri2012,agnihotri2013}. 
Symbols~$\rhd$~\cite{bug2017}, $\lhd$~\cite{wolf2019}, and 
$\medstar$~\cite{fuss2009} for electron impact with equivelocity 
conversion.}
\label{fig:crossDNA_2}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Scaling rule}
\label{subsec:scaling}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Toburen rule}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

%\documentclass[pra,showpacs,twocolumn]{revtex4}
\documentclass[preprint,showpacs]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{gensymb}
\usepackage{float}
\usepackage{MnSymbol}
\usepackage{gensymb}

@@

\author{A. M. P. Mendez, C. C. Montanari, J. E. Miraglia}
%\author{author}
%\author{author}
\address{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio 
(CONICET--UBA), \\ Buenos Aires, Argentina.}

\date{\today}% It is always \today, today,
\pacs{34.50Gb, 34.80Gs, 34.80Dp}

\begin{abstract}
In the present work, we investigate the ionization of molecules of 
biological interest by the impact of multicharged ions in the 
intermediate to high energy range. We performed full non--perturbative 
distorted--wave calculations (CDW) for thirty--six collisional systems. 
Six atomic targets compose these systems: H, C, N, O, F and S --which 
are the constituents of most of the DNA and biological molecules-- and 
six charged projectiles (antiprotons, H, He, B, C, and O). On account 
of the radiation damage caused by secondary electrons, we also inspect 
the energy and angular distributions of the emitted electrons from the 
atomic targets. We examine seventeen molecules: DNA and RNA bases, 
DNA backbone, pyrimidines, tetrahydrofuran (THF), and several 
C$_n$H$_n$ compounds. We show that the simple stoichiometric model 
(SSM), which approximates the molecular ionization cross sections as a 
linear combination of the atomic ones, gives reasonably good results 
for complex molecules. We also inspect the extensively used Toburen 
scaling of the total ionization cross sections of molecules with the 
number of weakly bound electrons. Based on the present atomic CDW 
results, we propose new active electron numbers. The CDW--based scaling 
improves significantly the single curve behavior for all the targets 
and ions studied here, from the intermediate to the high energy region. 
Furthermore, the new scaling describes well the available experimental 
data for proton impact, including small molecules such as H$_2$, water, 
methane, and ammonia. For completeness, we perform full molecular 
calculations for five nucleobases and test a modified stoichiometric 
formula based on the Mulliken charge of the composite atoms. The 
difference introduced by the new stoichiometric formula is less than 
$3\%$, which indicates the reliability and robustness of the SSM to 
deal with this type of molecules. The results of the extensive 
ion--target examination included in the present study allow us to 
assert that the SSM and the CDW--based scaling will be useful tools in 
this area.
\end{abstract}

@@

To overcome the first perturbative approximation limitations, and since 
the projectiles are multicharged ions, we resort to the Continuum 
Distorted Wave--Eikonal Initial State 
(CDW)~\cite{fainstein1988,miraglia2008,miraglia2009}, which includes higher 
perturbative corrections. Details on the CDW calculation are given in 
Section~\ref{sec:atoms}.
We start from the premise that the ionization 
process is the mechanism that deposits the most significant amount of 
primary energy. Moreover, the residual electrons from the ionization 
are known to be a source of significant local biological damage. 
The secondary electrons are included in Monte Carlo simulations, 
and hence their behavior requires further investigation.
In Section~\ref{subsec:meanener} and \ref{subsec:meanang}, we calculate
the mean energy and angular distributions of the ejected electrons. 
Surprisingly, we found a substantial dependence on the projectile charge, 
which is unexpected in the first Born approximation.

In Section~\ref{subsec:stoichiometric}, we deal with the molecular 
structure complexity of the targets by 
implementing the simplest stoichiometric model (SSM): the molecules are 
assumed to be composed of isolated independent atoms, and the total 
cross section by a linear combination of stoichiometric weighted atomic 
calculations.
By implementing the CDW and the SSM, we calculate ionization 
cross section of several molecules of biological interest, including 
DNA and RNA molecules, such as adenine, cytosine, guanine, thymine, uracil, 
tetrahydrofuran~(THF), pyrimidine, and DNA backbone, by the impact 
of antiprotons, H$^{+}$, He$^{+2}$, Be$^{+4}$, C$^{+6}$, and O$^{+8}$. 
In Section~\ref{subsec:scaling}, we
test the Toburen scaling rule~\cite{toburen1975,toburen1976}, which 
states that the ratio between the ionization cross section and the 
number of weakly bound electrons can be arranged in a narrow universal 
band in terms of the projectile velocity. We applied this rule to 
several hydrocarbons and nucleobases and noted that the width of the 
resulting universal band could be significantly reduced if we consider 
the number of active electrons in the collision based on the 
CDW results for the different atoms. The new scaling was then tested 
theoretically and by comparison with experimental data available.

The approach SSM considers the atoms in the molecule as neutral, 
which is not correct. In Section~\ref{subsec:molcalculations}, 
we used the molecular electronic structure code 
{\sc gamess}~\cite{gamess} to calculate the excess or defect of 
electron density on the atoms composing the molecules. Then, we modified 
the SSM to account for the departure from the neutrality of the atoms. 
We find that the modified SSM for the DNA molecules does not introduce 
substantial changes in the cross sections.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory: Ionization of Atoms}
\label{sec:atoms}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In the present study, we consider six atoms, $\alpha=$ H, C, N, O, P, 
and S, and six projectiles, antiprotons $\bar{p}$, H$^{+}$, He$^{+2}$, 
Be$^{+4}$, C$^{+6}$, and O$^{+8}$. 
Most of the organic molecules are composed of these atoms. Some 
particular molecules also include halogen atoms such as fluor and 
bromine; ionization cross sections of these elements have been 
previously published~\cite{miraglia2008}.

The total ionization cross sections of these atoms were calculated using 
the CDW. The initial bound and final continuum radial wave 
functions were obtained by using the {\sc radialf} code, developed by 
Salvat and co--workers~\cite{salvat1995}, and a Hartree--Fock potential 
obtained from the Depurated Inversion Method~\cite{mendez2016,mendez2018}. 
We used a few thousand pivot points to solve the Schr\"{o}dinger 
equation, depending on the number of oscillations of the continuum 
state. The radial integration was performed using the cubic spline 
technique. We expand our final continuum wave function as usual,
\begin{equation}
\psi_{\overrightarrow{k}}^{-}(\overrightarrow{r})=\sum_{l=0}^{l_{\max
}}\sum_{m=-l}^{l}R_{kl}^{-}(r)Y_{l}^{m}(\widehat{r})Y_{l}^{m^{\ast }}
(\widehat{k})\,.
\label{eq:contwave}
\end{equation}
The number of angular momenta considered varied from 8, at 
very low ejected--electron energies, up to $l_{\max}\sim 30$, 
for the highest energies 
considered. The same number of azimuth angles were required to obtain 
the four--fold differential cross section. The calculation performed does 
not display prior--post discrepancies at all. Each atomic total cross 
section was calculated using 35 to 100 momentum transfer values, 28 
fixed electron angles, and around 45 electron energies depending on the 
projectile impact energy. 
Further details of the calculation are given in Ref.~\cite{montanari2017}.

\begin{figure*}[t!]
\centering
%\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/Fig1.eps}
\includegraphics[width=0.8\textwidth]{figuras/atomicscaling.eps}
\caption{Reduced CDW total ionization cross section of six atomic 
targets. The curves are labeled with the charge state corresponding to 
the six multicharged projectiles.}
\label{fig:atomscaling}
\end{figure*}

We display our total CDW ionization cross sections for the six essential 
elements by the impact of the six projectiles in Fig.~\ref{fig:atomscaling}.
To reduce the resulting 36 magnitudes into a single consistent 
figure, we considered the fact that in the first Born approximation
the ionization cross section scales with the square of the projectile 
charge, $Z^{2}$. The impact energies considered 
range between 0.1 to 10 MeV/amu, where the CDW is supposed 
to hold. In fact, for the highest projectile charges the minimum 
impact energy where the CDW is expected to be valid could be 
higher than 100 keV. We also performed similar calculations with the 
first Born approximation, and we corroborated that it provides quite 
reliable results only for energies higher than a couple of MeV/amu. 
We use the same line color to indicate the projectile charge throughout 
all the figures of this work: dashed--red, solid--red, blue, magenta, 
olive and orange for antiprotons, H$^{+}$, He$^{+2}$, Be$^{+4}$, 
C$^{+6}$, and O$^{+8}$, respectively. Notably, there is no complete 
tabulation of ionization of atoms by the impact of multicharged ions. 
We hope that the ones presented in this article will be of help for 
future works.

Simultaneously, we will be reporting state to state ionization cross 
sections for the 36 ion--target systems considered in the present 
work~\cite{miraglia2019}. A great numerical effort was paid to obtain 
these results, and we expect that they will be useful to estimate 
molecule fragmentation.

@@

\begin{figure*}[t!]
\centering
%\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/fig5.eps}
\includegraphics[width=0.8\textwidth]{figuras/ener_mean.eps}
\caption{Mean emitted energy distribution for ionization by the impact 
of multicharged ions, given by Eq.~(\ref{40}). Dashed lines for $\bar{p}$,
solid lines for ion charges $+1$, $+2$, $+4$, $+6$ and $+8$,
as indicated.}
\label{fig:emittedener}
\end{figure*}

@@

In Fig.~\ref{fig:emittedener}, we estimate $\overline{E}_{\alpha}$ of
the emitted electron in the 0.5--2.7 a.u. energy range, or equivalently 
from 15 to 70 eV, for all the targets. Our results agree with the 
experimental findings~\cite{surdutovic2018}. As can be noted in the 
figure, the mean energy value is surprisingly sensitive to the 
projectile charge $Z$, which can duplicate the proton results in the 
intermediate region, i.e., 100--400 keV/amu. The effect observed can be 
attributed to the depletion caused by the multicharged ions to the 
yields of low energy electrons. This behavior cannot be found in the 
first Born approximation, where the $Z^2$ law cancels the $Z$ dependence
in Eq.~(\ref{40}). At high energies, $\overline{E}_{\alpha}$ tends to a 
universal value for all ions, as can be seen in Fig.~\ref{fig:emittedener}.

@@

As mentioned before, secondary electrons contribute to the total damage. 
Then, not only the ejection energy is essential but also the angle 
of emission. Once again, we can consider the single differential cross 
section in terms of the ejected electron solid angle $\Omega$, 
$d\sigma_{\alpha,nl}/d\Omega$, to be expressed as a distribution function, 
and the mean angle $\overline{\theta}_{\alpha}$ can be defined as
\begin{equation}
\overline{\theta}_{\alpha}=\frac{\langle\theta_{\alpha}\rangle}
{\langle 1\rangle}=\frac{1}{\sigma_{\alpha}}\sum\limits_{nl}
\int d\Omega\,\theta\,\frac{d\sigma_{\alpha,nl}}{d\Omega}
\end{equation}

\begin{figure*}[t!]
\centering
%\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/fig6.eps}
\includegraphics[width=0.8\textwidth]{figuras/ang_mean.eps}
\caption{Mean emitted angle distribution for ionization by impact of
multicharged ions.}
\label{fig:emittedang}
\end{figure*}

The mean emitted electron angles $\overline{\theta}_{\alpha}$ for the 
six atoms and six ions of interest are shown in Fig.~\ref{fig:emittedang}.
An significant dependence of $\overline{\theta}_{\alpha}$ with $Z$ is 
noticed for all the cases. Once again, this effect could not be 
observed in the first Born approximation (dotted line).

It is a general belief~\cite{Rudd1992} that the 
angular dispersion of emitted electrons is nearly isotropic. This 
behavior is caused by the insignificant angular anisotropy of 
sub--50--eV yield. 
A typical value for the ejection angle considered in the literature is 
$\overline{\theta}_{\alpha}\sim 70\degree$~\cite{surdutovic2018}, and 
it is quite correct in the range of validity of the first Born 
approximation for any target. However, when a distorted wave approximation 
is used, $\overline{\theta}_{\alpha}$ decreases substantially with $Z$ 
in the intermediate energy region. This effect is evident in 
Fig.~\ref{fig:emittedang};
for example, at 0.3 MeV/amu --where the Bragg peak for C$^{+6}$ impact 
occurs-- the CDW $\overline{\theta}_{\alpha}$ is half of the first Born 
value. This correction closes the damage to the forward direction.
We can attribute this forward direction correction to the capture to 
the continuum effect; the higher the charge $Z$, the smaller 
$\overline{\theta}$ will be. Of course, this effect only holds at 
intermediate energies and not at high impact energies.

Furthermore, Fig.~\ref{fig:emittedang} provides an illustrative 
description of the behavior of antiprotons: the projectile repels the 
electrons, being $\overline{\theta}_{\alpha}\sim 90\degree$. Note the 
opposite effect of proton and antiprotons respect to the first Born 
approximation; this phenomenon constitutes an angular Barkas effect.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Ionization of Molecules}
\label{sec:molecules}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The stoichiometric model}
\label{subsec:stoichiometric}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Lets us consider a molecule $M$ composed by $n_{\alpha}$ atoms of the
element $\alpha$, the SSM approaches the total ionization cross section 
of the molecule $\sigma_{M}$ as a sum of ionization cross sections of 
the isolated atoms $\sigma_{\alpha}$ weighted by $n_{\alpha}$, 
\begin{equation}
 \sigma_{M}=\sum\limits_{\alpha}n_{\alpha}\sigma_{\alpha}\,.  
 \label{eq:sumion}
\end{equation}
We classified the molecular targets of our interest in three families: 
CH, CHN, and DNA, as in Table~\ref{tab:families}.

@@

In Fig.~\ref{fig:crossDNA_1}, we report the total ionization cross 
sections by the impact of multicharged ions for adenine, cytosine, 
guanine, and thymine, by combining the SSM and CDW results in 
Eq.~\ref{eq:sumion}. For adenine, the agreement with the experimental 
data available for proton impact~\cite{iriki2011} is excellent. To the 
best of our knowledge, there are no experimental data on ion--collision 
ionization for the rest of the molecules. We have also included in this 
figure electron impact measurements~\cite{rahman2016} with the 
corresponding equivelocity conversion for electron incident energies 
higher than 300~eV. In this region, the proton and electron cross 
section should converge. Although the electron impact measurements are 
above our findings for all the molecular targets, it is worth stating 
that our results agree very well with other electron impact theoretical 
predictions~\cite{mozejko2003,tan2018}.

\begin{figure*}[t!]
\centering
%\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/fig2.eps}
\includegraphics[width=0.8\textwidth]{figuras/adn1.eps}
\caption{Reduced CDW ionization cross section by the impact of 
multicharged ions. Experiments: \mbox{\Large$\circ$}~\cite{iriki2011} 
for proton impact and $\square$~\cite{rahman2016} for electron impact 
with equivelocity conversion.}
\label{fig:crossDNA_1}
\end{figure*}

Our results for uracil, DNA backbone, pyrimidine, and THF are displayed 
in Fig.~\ref{fig:crossDNA_2}. For uracil, the agreement with the 
experimental proton impact measurements by 
Itoh~{\it et al.}~\cite{itoh2013} is good.
However, for the same target, our theory is a factor of two above 
the experimental ionization measurements by Tribedi and 
collaborators~\cite{agnihotri2012,agnihotri2013} by the impact of 
multicharged ions.
Nonetheless, it should be stated that our theoretical results coincide 
with calculations by Champion, Rivarola, and 
collaborators~\cite{agnihotri2012,champion2012}, which may indicate a 
possible misstep of the experiments.

For pyrimidine, we show a comparison of our results with experimental 
data for proton impact by Wolff~\cite{wolff2014} and also for electron
impact ionization~\cite{bug2017} at high energies. 
The electron impact measurements 
agree with our calculations for energies higher than 500keV. 
Unexpectedly, the proton impact cross sections are significantly lower 
than our findings. 
Much more experiments are available for ionization of THF molecule by 
proton~\cite{wang2016} and by electron~\cite{bug2017,wolf2019,fuss2009} 
impact. Our SSM with CDW results show overall good agreement with these
data.

\begin{figure*}[t!]
\centering
%\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/fig3.eps}
\includegraphics[width=0.8\textwidth]{figuras/adn2.eps}
\caption{Reduced CDW ionization cross section by the impact of 
multicharged ions. Experiments: proton impact on 
$\triangle$ uracil~\cite{itoh2013}, 
$\bigtriangledown$ pyrimidine~\cite{wolff2014} and $\meddiamond$
THF~\cite{wang2016}. Impact of $\ominus$ C$^{+4}$, 
$\boxminus$ C$^{+6}$, O$^{+6}$, F$^{+6}$, and
$\triangle\mkern-14mu-$ O$^{+8}$, F$^{+8}$ on 
uracil~\cite{agnihotri2012,agnihotri2013}. 
Symbols~$\rhd$~\cite{bug2017}, $\lhd$~\cite{wolf2019}, and 
$\medstar$~\cite{fuss2009} for electron impact with equivelocity 
conversion.}
\label{fig:crossDNA_2}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Scaling rules}
\label{subsec:scaling}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Toburen rule}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
